---
title: "Analysis For Camille"
format: html
code-fold: true
code-summary: "Expand this to see code"
---

Questions to answer:

* Which districts by type have the biggest discrepancy between cost of living and their average salary?
* What proportion of districts in each category that are above/below alice threshold?
* How many districts of each type are under 5000 enrollment? What percentage of districts in each type are under 5000 enrollment?
*  How many urban/suburban districts are under 5000 and how do they pay their teachers?
* What number of districts in Texas pay their teachers under living wage (or percentage)?
* What percentage of students are served by teachers that are underpaid?
* Are there more districts being helped in the TEA "Rural" definition or the bill's enrollment breakdown?

NOTE: I used 2024 data to answer these, but can easily look at it for another year or overtime. 

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(janitor)
library(scales)
library(sf)
library(mapview)
library(RColorBrewer)

```


Here's a sample of the data.

```{r}
#| label: import
#| message: false
#| warning: false
#| echo: false

above_below_alice <- read_rds("data-processed/above_below_alice.rds")

above_below_alice |> head()
```

## Which district types have the biggest discrepancy between cost of living and average salary in 2024?

### By TEA district type

```{r}
above_below_alice |> filter(end_year == 2024) |> 
  group_by (tea_description) |> 
  summarize( avg_difference = mean(difference),
             total_districts_of_this_type = n(),
             total_students_affected = sum(total_enrollment)) |> 
  arrange(avg_difference)
```

Of the TEA district types, Non-metropolitan Fast Growing districts have the worst discrepancy between the average salary for a district and that district's county's ALICE cost of living. There are 34 districts and almost 62,000 students that fall into this category of Non-metropolitan fast-growing.

### By bill category (enrollment)

```{r}
above_below_alice |> filter(end_year == 2024) |> 
  group_by (bill_category) |> 
  summarize( avg_difference = mean(difference),
             total_districts_of_this_type = n(),
             total_students_affected = sum(total_enrollment)) |> 
  arrange(avg_difference)
```

Of the categories outlined by the bill (enrollment-based), districts that have under 1600 students (but more than 300) have the worst discrepancy between the average salary for a district and that district's county's ALICE cost of living. There are 425 districts and about 340,000 students that fall into this category of under 1600 enrollment. 

## What proportion of districts in each category are above/below alice threshold in 2024?

### Based on the TEA categories

```{r}
above_below_alice |> filter(end_year == 2024) |> 
  group_by(tea_description) |> 
  summarize(under_alice = sum(above_alice == FALSE),
            total_districts_2024 = n()) |> 
  mutate(prop_under_alice = (under_alice / total_districts_2024) |> round_half_up(3)) |> 
  arrange(desc(prop_under_alice))
```
The above table shows the number of districts in each TEA district type category where the average salary does not meet the ALICE cost of living threshold (under_alice), the total number of districts in 2024 that fell into that district type category (total_districts_2024) and the proportion of districts in that district type category that where the average salary does not meet the ALICE cost of living threshold (prop_under_alice).

### Based on the bill categories (enrollment)

```{r}
above_below_alice |> filter(end_year == 2024) |> 
  group_by(bill_category) |> 
  summarize(under_alice = sum(above_alice == FALSE),
            total_districts_2024 = n()) |> 
  mutate(prop_under_alice = (under_alice / total_districts_2024) |> round_half_up(3)) |> 
  arrange(desc(prop_under_alice))
  
```

This table shows the number of districts in each bill-defined enrollment category where the average salary does not meet the ALICE cost of living threshold (under_alice), the total number of districts in 2024 that fell into that bill-defined enrollment category (total_districts_2024) and the proportion of districts in that category that where the average salary does not meet the ALICE cost of living threshold (prop_under_alice).

## How many districts of each type are under 5000 enrollment and what proportion of districts in each type are under 5000 enrollment?

This is based on the TEA district types. 

```{r}
above_below_alice |> filter(end_year == 2024 ) |> 
  group_by(tea_description) |> 
  summarize(districts_under_5000 = sum(total_enrollment < 5000),
            total_districts = n()) |> 
  mutate(
    prop_districts_under_5000 = (districts_under_5000 / total_districts) |> round_half_up(3)
  ) |> arrange(desc(prop_districts_under_5000))
```

In 2024, the "Rural" district type had the most districts with enrollment under 5000 at 461. 

## How many major urban/suburban districts are under 5000 and how do they pay their teachers?

Based on the table from the last question, there were 0 districts in 2024 that were considered Major Urban and had enrollment under 5000. There were 7 districts in 2025 that were considered Major Suburban and had enrollment under 5000. Let's look at those 7. 

```{r}
above_below_alice |> filter(tea_description == "Major Suburban" & end_year == 2024 & total_enrollment < 5000) # add select to simplify
```

Of these 7 districts, 3 had an average salary that did not meet the ALICE cost of living threshold.

## What number of districts in Texas pay their teachers under living wage? What's that in a percentage?

```{r}
above_below_alice |> ungroup() |> filter(end_year == 2024 & above_alice == F) |> 
  summarize(total_districts_in_Texas_under_alice_threshold = n())
```

In 2024, 463 districts in Texas on average paid their teachers under a living wage.

Now, the percentage.

```{r}
above_below_alice |> ungroup() |>  filter(end_year == 2024) |> 
  summarise( under_alice = sum(above_alice == F),
             total_districts = n()) |> 
  mutate(percent_districts_under_alice_threshold = ((under_alice / total_districts)*100) |> round_half_up(1))
```

In 2024, 45.4% of districts in Texas on average paid their teachers under a living wage.

## What percentage of students are served by teachers that are underpaid?

```{r}
above_below_alice |> filter(end_year == 2024 ) |> ungroup() |> 
  summarise(total_students_served_by_underpaid_teachers = sum(total_enrollment[above_alice == F]),
    total_students = sum(total_enrollment)
  ) |> mutate(
    percent_students_served_by_underpaid_teachers = (total_students_served_by_underpaid_teachers / total_students * 100) |> round_half_up(2)
  ) #|> select(percent_students_served_by_underpaid_teachers)
```
In 2024, 31.15% of students were served by teachers who were on average paid under a living wage. 

## Are there more districts being helped in the TEA "Rural" definition or the bill's enrollment breakdown?

Districts being helped by TEA "Rural" description

```{r}
above_below_alice |> filter(end_year == 2024 & tea_description == "Rural") |> ungroup() |> 
  count(name = "total_districts_TEA_rural")
```

464 districts would get money if the bill targeted TEA's "Rural" districts based on 2022-2023 TEA district types. 

Distircts being helped based on bill categories of enrollment.

```{r}
above_below_alice |> filter(end_year == 2024 & (bill_category == "under 300" | bill_category == "under 1600" | bill_category == "under 5000")) |> 
  group_by(bill_category) |> 
  count(name = "total_districts") |> adorn_totals()
```

827 total districts would get money with the bill targeting districts under 5000, under 1600 and under 300 enrollment based on 2023-2024 enrollment. 

