---
title: "Analysis For Camille"
format: html
---

Questions?
- Which districts by type have the biggest discrepancy between cost of living and their average salary?
- What proportion of districts in each category that are above/below alice threshold?
- How many districts of each type are under 5000 enrollment? What percentage of districts in each type are under 5000 enrollment?
- How many urban/suburban districts are under 5000 and how do they pay their teachers?
- What number of districts in Texas pay their teachers under living wage (or percentage)?
- What percentage of students are served by teachers that are underpaid?
- Are there more districts being helped in the TEA "Rural" definition or the bill's enrollment breakdown?

NOTE: I used 2024 data to answer these, but can easily look at it for another year or overtime. 

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(scales)
library(sf)
library(mapview)
library(RColorBrewer)

```

## Import

```{r}
above_below_alice <- read_rds("data-processed/above_below_alice.rds")

above_below_alice |> head()
```

## Which district types have the biggest discrepancy between cost of living and average salary in 2024?

By TEA district type
```{r}
above_below_alice |> filter(end_year == 2024) |> 
  group_by (tea_description) |> 
  summarize( avg_difference = mean(difference),
             total_districts_of_this_type = n(),
             total_students_affected = sum(total_enrollment)) |> 
  arrange(avg_difference)
```
Of the TEA district types, Non-metropolitan Fast Growing districts have the worst discrepancy between the average salary for a district and that district's county's ALICE cost of living. There are 34 districts and almost 62,000 students that fall into this category of Non-metropolitan fast-growing.

By bill category (enrollment)
```{r}
above_below_alice |> filter(end_year == 2024) |> 
  group_by (bill_category) |> 
  summarize( avg_difference = mean(difference),
             total_districts_of_this_type = n(),
             total_students_affected = sum(total_enrollment)) |> 
  arrange(avg_difference)
```
Of the categories outlined by the bill (enrollment-based), districts that have under 1600 students (but more than 300) have the worst discrepancy between the average salary for a district and that district's county's ALICE cost of living. There are 425 districts and about 340,000 students that fall into this category of under 1600 enrollment. 

## What proportion of districts in each category that are above/below alice threshold in 2024?

First, based on the TEA categories.
```{r}
above_below_alice |> filter(end_year == 2024) |> 
  group_by(tea_description) |> 
  summarize(under_alice = sum(above_alice == FALSE),
            total_districts_2024 = n()) |> 
  mutate(prop_under_alice = (under_alice / total_districts_2024) |> round_half_up(3)) |> 
  arrange(desc(prop_under_alice))
```
The above table shows the number of districts in each TEA district type category where the average salary does not meet the ALICE cost of living threshold (under_alice), the total number of districts in 2024 that fell into that district type category (total_districts_2024) and the proportion of districts in that district type category that where the average salary does not meet the ALICE cost of living threshold (prop_under_alice).


Now, based on the bill categories (enrollment).
```{r}
above_below_alice |> filter(end_year == 2024) |> 
  group_by(bill_category) |> 
  summarize(under_alice = sum(above_alice == FALSE),
            total_districts_2024 = n()) |> 
  mutate(prop_under_alice = (under_alice / total_districts_2024) |> round_half_up(3)) |> 
  arrange(desc(prop_under_alice))
  
```

This table shows the number of districts in each bill-defined enrollment category where the average salary does not meet the ALICE cost of living threshold (under_alice), the total number of districts in 2024 that fell into that bill-defined enrollment category (total_districts_2024) and the proportion of districts in that category that where the average salary does not meet the ALICE cost of living threshold (prop_under_alice).

## How many districts of each type are under 5000 enrollment and what proportion of districts in each type are under 5000 enrollment?

This is based on the TEA district types. 
```{r}
above_below_alice |> filter(end_year == 2024 ) |> 
  group_by(tea_description) |> 
  summarize(districts_under_5000 = sum(total_enrollment < 5000),
            total_districts = n()) |> 
  mutate(
    prop_districts_under_5000 = districts_under_5000 / total_districts
  ) |> arrange(desc(prop_districts_under_5000))
```
In 2024, the "Rural" district type had the most districts with enrollment under 5000 at 461. 

## How many urban/suburban districts are under 5000 and how do they pay their teachers?

Based on the table from the last question, there were 0 districts in 2024 that were considered Major Urban and had enrollment under 5000. There were 7 districts in 2025 that were considered Major Suburban and had enrollment under 5000. Let's look at those 7. 

```{r}
above_below_alice |> filter(tea_description == "Major Suburban" & end_year == 2024 & total_enrollment < 5000)
```
Of these 7 districts, 3 had an average salary that did not meet the ALICE cost of living threshold.

## What number of districts in Texas pay their teachers under living wage? What's that in a percentage?


```{r}
above_below_alice |> filter(end_year == 2024 & above_alice == F) |> ungroup() |> 
  summarize(total_districts_in_Texas_under_alice_threshold = n())
```
In 2024, 463 districts in Texas on average paid their teachers under a living wage.

Now, the percentage.
```{r}
above_below_alice |> filter(end_year == 2024) |> ungroup() |> 
  summarise(percent_districts_under_alice_threshold = ((sum(above_alice == F) / n())*100) |> round_half_up(2))
```
In 2024, 45.39% of districts in Texas on average paid their teachers under a living wage.

## What percentage of students are served by teachers that are underpaid?

```{r}
above_below_alice |> filter(end_year == 2024 ) |> ungroup() |> 
  summarise(total_students = sum(total_enrollment),
            total_students_served_by_underpaid_teachers = sum(total_enrollment[above_alice == F])
  ) |> mutate(
    percent_students_served_by_underpaid_teachers = (total_students_served_by_underpaid_teachers / total_students * 100) |> round_half_up(2)
  ) |> select(percent_students_served_by_underpaid_teachers)
```
In 2024, 31.15% of students were served by teachers who were on average paid under a living wage. 
